---
title: "Poisson and Royston-Parmar comparison models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Poisson and Royston-Parmar comparison models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
library(estimandYLL)
```

The comparison functions standardize both exposure scenarios over the **full
analytic sample**. They do not contrast the two observed groups' separate
covariate distributions. Compare them with `estimate_yll(target_population =
"all")` to isolate differences due to the mortality model and estimation method.
A comparison with the exposed or unexposed target also changes the population.

## Poisson model

Follow-up is split by attained age using a Lexis expansion. The model includes
exposure, a natural spline of the interval midpoint, their interaction, baseline
covariates, and a log person-time offset. Predicted interval death counts are
converted to death probabilities as one minus their negative exponential.

```{r poisson}
data(yll_toy)
poisson <- estimate_yll_poisson(
  yll_toy[1:500, ], time_var = "period", event_var = "event",
  exposure_var = "hypertension", reference_level = "No", exposed_level = "Yes",
  age_at_entry_var = "age", confounders_baseline = "sex",
  age_start = 50, age_end = 70, age_interval = 10, B = 0
)
poisson$summary
```

## Royston-Parmar model

This optional method uses `rstpm2::stpm2()` with attained age and delayed entry.
The default has four degrees of freedom for the baseline log cumulative hazard
and proportional exposure and covariate effects. `rp_model = "stratified"`
fits separate models by exposure group; it changes the model specification.

```{r royston-parmar}
if (requireNamespace("rstpm2", quietly = TRUE)) {
  rp <- estimate_yll_royston_parmar(
    yll_toy[1:500, ], time_var = "period", event_var = "event",
    exposure_var = "hypertension", reference_level = "No", exposed_level = "Yes",
    age_at_entry_var = "age", confounders_baseline = "sex",
    age_start = 50, age_end = 70, age_interval = 10, B = 0
  )
  rp$summary
}
```

The manuscript did not report results for this model because its bootstrap
variance was unstable in that application. Its availability here is not an
assurance of stability on other data. Inspect model warnings, replicate estimates
and `meta$bootstrap_failures` before interpreting an interval. Finite but impossible
bootstrap ERLs are also flagged in `meta$bootstrap_unreliable`. Their raw values
are retained for diagnosis; intervals based on such a distribution are unreliable.

## Shared output and uncertainty

Both functions return the same five elements as the main estimator and use the
same direction, reference ERL minus exposed ERL. Each person's curve is conditioned
at the starting age before averaging. Both use normal-approximation participant
bootstrap intervals; they do not expose the main function's `ci_method` option.
Their default prediction interval is one year and can be changed with
`prediction_interval`. The reporting interval remains `age_interval`.
