---
title: "Getting started with estimandYLL"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with estimandYLL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

`estimandYLL` estimates **years of life lost (YLL)** and counterfactual
life expectancy under the parametric g-formula. The main interface is
`estimand_yll()`.

The estimand is stated through three arguments:

1. `target_population`: who to average over.
2. `intervention`: which exposure worlds or policy change to compare.
3. `measure`: how to display the result.

```{r}
library(estimandYLL)
data(yll_toy)
str(yll_toy)
```

## Full contrasts

`intervention = "full_contrast"` compares the all-reference and all-exposed
counterfactual worlds. The target population determines whether the result
is ATE-like, ATT-like, or ATC-like.

```{r ate, eval = FALSE}
res_all <- estimand_yll(
  data = yll_toy,
  target_population = "all",
  intervention = "full_contrast",
  measure = "yll",
  B = 50,
  method = "normal",
  show_progress = FALSE,
  id_var = "id",
  time_var = "period",
  event_var = "event",
  exposure_var = "hypertension",
  reference_level = "No",
  exposed_level = "Yes",
  age_at_entry_var = "age",
  age_start = 50,
  age_end = 90,
  age_interval = 5,
  confounders_baseline = c("sex", "education_years", "bmi", "smoke_binary"),
  use_future = FALSE
)

res_all$summary
```

Observed-exposed and observed-unexposed target populations are specified
directly:

```{r att-atc, eval = FALSE}
res_exposed <- estimand_yll(
  data = yll_toy,
  target_population = "exposed",
  intervention = "full_contrast",
  measure = "yll",
  B = 50, method = "normal", show_progress = FALSE, use_future = FALSE,
  id_var = "id", time_var = "period", event_var = "event",
  exposure_var = "hypertension",
  reference_level = "No", exposed_level = "Yes",
  age_at_entry_var = "age",
  age_start = 50, age_end = 90, age_interval = 5,
  confounders_baseline = c("sex", "education_years", "bmi", "smoke_binary")
)

res_unexposed <- estimand_yll(
  data = yll_toy,
  target_population = "unexposed",
  intervention = "full_contrast",
  measure = "life_year_change",
  B = 50, method = "normal", show_progress = FALSE, use_future = FALSE,
  id_var = "id", time_var = "period", event_var = "event",
  exposure_var = "hypertension",
  reference_level = "No", exposed_level = "Yes",
  age_at_entry_var = "age",
  age_start = 50, age_end = 90, age_interval = 5,
  confounders_baseline = c("sex", "education_years", "bmi", "smoke_binary")
)
```

With `measure = "life_year_change"`, an ATC-like harmful change is reported
as a negative value.

## Partial-change policies

`intervention = "partial_change"` compares the natural observed exposure
distribution with a policy-like change in which only a specified fraction
changes exposure state. In statistical terminology this is a stochastic
intervention, but the API uses clinical policy language.

Example: 30% of observed-exposed individuals move to the reference level,
averaged over observed-exposed individuals.

```{r partial-change, eval = FALSE}
res_quit <- estimand_yll(
  data = yll_toy,
  target_population = "exposed",
  intervention = "partial_change",
  change_from = "exposed",
  change_to = "reference",
  change_probability = 0.30,
  measure = "life_year_change",
  B = 50, show_progress = FALSE, use_future = FALSE,
  id_var = "id", time_var = "period", event_var = "event",
  exposure_var = "smoke_binary",
  reference_level = "Never", exposed_level = "Current/Ever",
  age_at_entry_var = "age",
  age_start = 50, age_end = 90, age_interval = 5,
  confounders_baseline = c("sex", "education_years", "bmi")
)
```

## Visualisation

```{r plots, eval = FALSE}
plot_yll_estimate(res_all)
plot_conditional_survival(res_all, age_start = 60)
plot_marginal_survival(res_all)
```

## Result object

`summary` contains one row per starting age and includes the selected
`estimate`, confidence interval, `target_population`, `intervention`, and
`measure`. `detailed_results` additionally keeps the original `yll`,
life-year change, and arm-specific life expectancies for checking.
